### 2018-08-25 conclusion.tex

Only the batch-wise online reprocessing approach has been treated in this 
work. However, the SERPENT 2 Monte Carlo code was recently extended for 
continuous online fuel reprocessing simulation \cite{aufiero_extended_2013}. 
This extension must be verified against existing SaltProc/SERPENT or 
ChemTriton/SCALE packages, and could be employed for simulating immediate removal of 
fission product gases (e.g., Xe, Kr) which have a strong negative impact on 
core lifetime and breeding efficiency. Finally, using the built-in SERPENT 2 
Monte Carlo code online reprocessing \& refueling material burnup routine would 
significantly speed up computer-intensive full-core depletion simulations.

Lastly, this work enables accident safety analysis such as multi-physics 
simulation of the \gls{MSBR} with the coupled 
neutronics/thermal-hydraulics code, Moltres \cite{lindsay_introduction_2018}. 
The existing full-core SERPENT 2 model and equilibrium fuel material 
composition would be employed to generate problem-oriented nuclear data 
libraries for further usage in accident transient analysis. The final goal of 
this effort is to develop a fast-running computational model which could study 
the dynamic behavior of generic \glspl{MSR}, performing detailed safety 
analysis and design optimization for a variety of reactor concepts.

>>>>

Only the batch-wise online reprocessing approach has been treated in this 
thesis. However, the SERPENT 2 Monte Carlo code was recently extended for 
continuous online fuel reprocessing simulation \cite{aufiero_extended_2013}. 
This extension must be verified against existing SaltProc/SERPENT or 
ChemTriton/SCALE packages, and could be employed for immediate removal of 
fission product gases (e.g., Xe, Kr) which have a strong negative impact on 
core lifetime and breeding efficiency. Finally, using the built-in SERPENT 2 
Monte Carlo code online reprocessing \& refueling material burnup routine would 
significantly speed up computer-intensive full-core depletion simulations.

### 2018-08-25 introduction.tex

\begin{table}[ht!]
\caption{Tools and methods for fast spectrum system fuel cycle analysis.}
\begin{tabularx}{\textwidth}{ x | s | s | m | x } \hline & Neutronic code  & 
Depletion code    & \qquad Authors & Spectra   \\
\hline
1 & \gls{MCNP} \cite{noauthor_mcnp_2004}      & REM 
\cite{heuer_simulation_2010}  & Doligez \emph{et al.}, 2014; Heuer \emph{et 
al.}, 2014  \cite{doligez_coupled_2014,heuer_towards_2014}    & fast \\
\hline
2 & ERANOS \cite{ruggieri_eranos_2006}      & ERANOS     & Fiorina \emph{et 
al.}, 2013 \cite{fiorina_investigation_2013}            & fast \\
\hline
3 & KENO-IV \cite{goluoglu_monte_2011}     & ORIGEN \cite{gauld_isotopic_2011}     
& Sheu \emph{et al.}, 2013 \cite{sheu_depletion_2013} & fast \\
\hline
4 & SERPENT 2 \cite{leppanen_serpent_2015-1}   & SERPENT 2  & Aufiero \emph{et 
al.}, 2013 \cite{aufiero_extended_2013} & fast \\
\hline
5 & MCODE \cite{xu_mcode_2008}      & ORIGEN2 \cite{croff_users_1980}      & 
Ahmad \emph{et al.}, 2015 \cite{ahmad_neutronics_2015}   & thermal  \\
\hline
6 & \gls{MCNP}6     & CINDER90 \cite{goorley_mcnp6_2013}     & Park \emph{et 
al.}, 2015; Jeong \emph{et al.}, 2016 \cite{park_whole_2015, 
jeong_equilibrium_2016}& thermal\\
\hline
7 & SCALE \cite{bowman_scale_2011}      & SCALE/ ChemTriton 
\cite{powers_new_2013}    & Powers \emph{et al.}, 2014; Betzler \emph{et al.}, 
2017 \cite{powers_new_2013,powers_inventory_2014,betzler_molten_2017}& 
thermal\\
\hline
8 & SERPENT 2      & SERPENT 2     & Rykhlevskii \emph{et al.}, 2017 
\cite{rykhlevskii_online_2017} & thermal\\
\hline
9 & \gls{MCNP}      & REM  & Nuttin \emph{et al.} 
\cite{nuttin_potential_2005}&thermal  \\ \hline
\end{tabularx}
  \label{tab:fs_codes}
\end{table}
\FloatBarrier

@@

Methods (1, 3, 4) simulate some form of reactivity control, and methods (1, 4, 
5, 6, 8, 9) use a set of all nuclides in depletion calculations.

Many liquid-fueled \gls{MSR} designs rely on online fuel processing in which 
material moves to and from the core at all times (continuous) or at specific 
time steps (batch). To account for batch discharge, a depletion tool must 
remove some or all material at specified intervals. This requires the burn-up 
simulation to stop at a given time and restart with a new liquid fuel 
composition (after removal of discarded materials and addition of 
fissile/fertile materials). Accounting for continuous removal or addition  
presents a greater challenge since it requires adding a term to the Bateman 
equations. In SCALE \cite{bowman_scale_2011}, ORIGEN \cite{gauld_isotopic_2011} 
solves a set of Bateman equations using spectrum-averaged fluxes and cross 
sections generated from a deterministic transport calculation. Methods (1, 4, 
8) model true continuous feeds and removals, while other methods employ a 
batch-wise approach. \gls{ORNL} researchers have developed ChemTriton, a 
Python-based script for SCALE/TRITON which uses a semi-continuous batch process 
to simulate a continuous reprocessing. ChemTriton models salt treatment, 
separations, discharge, and refill using a unit-cell \gls{MSR} SCALE/TRITON 
model over small time steps to simulate continuous reprocessing and deplete the 
fuel salt \cite{powers_new_2013}.

Thorium-fueled \gls{MSBR}-like reactors similar to the one in this work are 
described in (6, 7, 8, 9). Nevertheless, most of these efforts considered only 
simplified unit-cell geometry because depletion computations for a many-year 
fuel cycle are computationally expensive even for simple models.

>>>>

\begin{table}[ht!]
\caption{Tools and methods for fast spectrum system fuel cycle analysis.}
\begin{tabularx}{\textwidth}{ m | b | x } 
\hline Neutronic code    & \qquad Authors & Spectra   \\
\hline
\gls{MCNP}/REM \cite{noauthor_mcnp_2004,heuer_simulation_2010}  & Doligez \emph{et al.}, 2014; Heuer \emph{et al.}, 2014  \cite{doligez_coupled_2014,heuer_towards_2014}    & fast \\
\hline
ERANOS \cite{ruggieri_eranos_2006}  & Fiorina \emph{et al.}, 2013 \cite{fiorina_investigation_2013}            & fast \\
\hline
KENO-IV/ORIGEN \cite{goluoglu_monte_2011,gauld_isotopic_2011}     & Sheu \emph{et al.}, 2013 \cite{sheu_depletion_2013} & fast \\
\hline
SERPENT 2 \cite{leppanen_serpent_2015-1}  & Aufiero \emph{et al.}, 2013 \cite{aufiero_extended_2013} & fast \\
\hline
MCODE/ORIGEN2 \cite{xu_mcode_2008,croff_users_1980}      & Ahmad \emph{et al.}, 2015 \cite{ahmad_neutronics_2015}   & thermal  \\
\hline
\gls{MCNP}6/CINDER90 \cite{goorley_mcnp6_2013}     & Park \emph{et al.}, 2015; Jeong \emph{et al.}, 2016 \cite{park_whole_2015, jeong_equilibrium_2016}& thermal\\
\hline
SCALE/TRITON \cite{bowman_scale_2011,powers_new_2013}    & Powers \emph{et al.}, 2014; Betzler \emph{et al.}, 2017 \cite{powers_new_2013,powers_inventory_2014,betzler_molten_2017}& thermal\\
\hline
SERPENT 2     & Rykhlevskii \emph{et al.}, 2017 \cite{rykhlevskii_online_2017} & thermal\\
\hline
\gls{MCNP}/REM  & Nuttin \emph{et al.} \cite{nuttin_potential_2005}&thermal  \\ \hline
\end{tabularx}
  \label{tab:fs_codes}
\end{table}
\FloatBarrier

@@

References \cite{doligez_coupled_2014,heuer_towards_2014,sheu_depletion_2013,
aufiero_extended_2013} provide some form of reactivity control, and methods 
\cite{doligez_coupled_2014,heuer_towards_2014,aufiero_extended_2013,
ahmad_neutronics_2015, park_whole_2015, jeong_equilibrium_2016,rykhlevskii_online_2017,
nuttin_potential_2005} use a set of all nuclides in depletion calculations.

Liquid-fueled \gls{MSR} designs have online separations and feeds, where material 
is moved to or from the core at specific time steps (batch) or continuously. In 
batch-wise approach the burn-up simulation stops at a given time and restart with
a new liquid fuel composition (after removal of discarded materials and addition 
of fissile/fertile materials). \gls{ORNL} researchers have developed ChemTriton, 
a Python-based script for SCALE/TRITON which the batch-wise to simulate a continuous 
reprocessing. This tool models salt treatment, separations, discharge, and refill 
using a unit-cell \gls{MSR} SCALE/TRITON depletion simulation over small time steps 
to simulate continuous reprocessing and deplete the fuel salt \cite{powers_new_2013}. 
Methods listed in references \cite{fiorina_investigation_2013,sheu_depletion_2013,
park_whole_2015, jeong_equilibrium_2016,powers_inventory_2014,betzler_molten_2017,
rykhlevskii_online_2017} and current work employ a batch-wise approach.

Accounting for a continuous removal or addition is more difficult because it requires
adding a term to the Bateman equations. In SCALE \cite{bowman_scale_2011}, 
ORIGEN \cite{gauld_isotopic_2011} solves a set of Bateman equations using 
spectrum-averaged fluxes and cross sections generated from a deterministic transport 
calculation. Approaches listed in references \cite{doligez_coupled_2014,heuer_towards_2014,
aufiero_extended_2013,nuttin_potential_2005} model true continuous feeds and removals.

Thorium-fueled \gls{MSBR}-like reactors similar to the one in this work are described in 
\cite{park_whole_2015, jeong_equilibrium_2016,powers_new_2013,powers_inventory_2014, 
betzler_molten_2017,rykhlevskii_online_2017,nuttin_potential_2005}. Nevertheless, most of 
these efforts considered only simplified unit-cell geometry because depletion computations 
for a many-year fuel cycle are computationally expesive even for simple models.

### 2018-08-25 methods.tex

Liquid-fueled system modeling with contemporary reactor physics codes is 
difficult because most of these codes were developed for analyzing solid-fueled 
reactors. In fact, world-wide commercial reactors fleet today is represented by 
systems with solid fuel. Liquid-fueled systems ability to continuously remove 
fission products and add fissile and/or fertile elements is sthe main challenge 
for depletion simulations. SaltProc takes into account online separations and 
feeds using the SERPENT 2 continuous-energy Monte Carlo neutron transport and 
depletion code.  All calculations presented in this study were performed using 
SERPENT 2 version 2.1.30 on Blue Waters’ XE6 nodes.

>>>>

Liquid-fueled systems ability to continuously remove fission products and add 
fissile and/or fertile elements is the main challenge for depletion simulations. 
Introduced in this work Python-package, SaltProc, takes into account online 
separations and feeds using SERPENT 2 continuous-energy Monte Carlo neutron 
transport and depletion code.

### 2018-08-25 results.tex

\section{Results}
The SaltProc online reprocessing simulation code is demonstrated in four applications: (1) analyzing  \gls{MSBR} neutronics and fuel cycle to find the equilibrium core composition and core depletion, (2) studying operational and safety parameters evolution during \gls{MSBR} operation, (3) demonstrating that in a single-fluid two-region \gls{MSBR} conceptual design the undermoderated outer core zone II works as a virtual ``blanket", reduces neutron leakage and improves breeding ratio due to neutron energy spectral shift, and (4) determining the effect of fission product removal on the core neutronics.

The neutron population per cycle and the number of active/inactive cycles were chosen to obtain balance between reasonable uncertainty for a transport problem ($\leq$ 15 pcm\footnote{ 1 pcm = 10$^{-5}\Delta k_{eff}/k_{eff}$} for effective multiplication factor) and computational time. The \gls{MSBR} depletion and safety parameter computations were performed on 64 Blue Waters XK7 nodes (two AMD 6276 Interlagos CPU per node, 16 floating-point Bulldozer core units per node or 32 ``integer" cores per node, nominal clock speed is 2.45 GHz). The total computational time for achieving equilibrium composition was approximately 9,900 node-hours (18 core-years.)

\subsection{Effective multiplication factor}
Figure~\ref{fig:keff} demonstrates the effective multiplication factors obtained using SaltProc and SERPENT 2. The effective multiplication factors are calculated after removing fission products listed in Table~\ref{tab:reprocessing_list} and adding the fertile material at the end of ``cycle time"\footnote{ The \gls{MSBR} program defined a ``cycle time" as the amount of time required to remove 100\% of a target nuclide from a fuel salt.} which was fixed at 3 days for this work. The effective multiplication factor fluctuates significantly as a result of the batch-wise nature of this online reprocessing strategy.

First, SERPENT calculates the effective multiplication factor for the beginning of cycle time (fresh fuel composition for the first step). Next, it computes the new fuel salt composition for the end of a 3-day depletion step. The corresponding effective multiplication factor is much smaller than the previous one. Finally, SERPENT calculates $k_{eff}$ for the depleted composition after applying feeds and removals, and this increases accordingly since major reactor poisons (e.g. Xe, Kr) are removed, while fresh fissile material ($^{233}$U) from the protactinium decay tank is added. 
\begin{figure}[ht!] % replace 't' with 'b' to 
  \centering
  \includegraphics[width=\textwidth]{keff.png}
  \caption{Effective multiplication factor dynamics for full-core \gls{MSBR} model for a 60-year reactor operation. The confidence interval $\sigma=\pm15pcm$ is shaded.}
  \label{fig:keff}
\end{figure}
Additionally, the presence of rubidium, strontium, cesium, and barium in the core are disadvantageous to reactor physics. In fact, removal of these elements every 3435 days causes the multiplication factor to jump by approximately 450 pcm, and limits using the batch approach for online reprocessing simulation. Overall, the effective multiplication factor gradually decreases from 1.075 to $k_{eff} \approx 1.02$ at equilibrium after approximately 6 years of irradiation.

\subsection{Fuel salt composition dynamics}
The analysis of the fuel salt composition evolution provides more comprehensive information about the equilibrium state. Figure~\ref{fig:adens_eq} shows number density of major nuclides which have a strong influence on the reactor core physics. Concentration of $^{233}$U, $^{232}$Th, $^{233}$Pa, and $^{232}$Pa in fuel salt change insignificantly after approximately 2500 days of operation. Particularly, $^{233}$U number density fluctuates less than 0.8\% in the time interval from 16 to 20 years of operation, hence,a quasi-equilibrium state was achieved after 16 years of reactor operation.
\begin{figure}[ht!] % replace 't' with 'b' to 
  \centering
  \includegraphics[width=\textwidth]{major_isotopes_adens.png}
  \caption{Number density of major nuclides during 60 years of reactor operation.}
  \label{fig:adens_eq}
\end{figure}
In contrast, a wide variety of nuclides, including fissile isotopes (e.g. $^{235}$U) and non-fissile strong absorbers (e.g. $^{234}$U), keep accumulating in the core. Figure~\ref{fig:fissile_short} demonstrate production of fissile isotopes in the core. In the end of considered operational time the core contains significant $^{235}$U ($\approx10^{-5}$ atom/b-cm), $^{239}$Pu ($\approx5\times10^{-7}$ atom/b-cm), and $^{241}$Pu ($\approx 5\times10^{-7}$ atom/b-cm). Meanwhile, the equilibrium number density of the target fissile isotope $^{233}$U was approximately 7.97$\times10^{-5}$ atom/b-cm. Thus, production of new fissile materials in the core as well as $^{233}$U breeding make it possible to compensate for negative effects of strong absorber accumulation and keep the reactor critical.
\begin{figure}[htp!] % replace 't' with 'b' to 
  \centering
  \includegraphics[width=\textwidth]{fissile_short.png}
  \caption{Number density of fissile in epithermal spectrum nuclides accumulation during the reactor operation.}
  \label{fig:fissile_short}
\end{figure}

\subsection{Neutron spectrum}
Figure~\ref{fig:spectrum} shows the normalized neutron flux spectrum for the full-core \gls{MSBR} model in the energy range from $10^{-8}$ to $10$ MeV. The neutron energy spectrum at equilibrium is harder than at startup due to $^{238}$Pu, $^{239}$Pu, $^{240}$Pu, $^{241}$Pu, and $^{242}$Pu accumulation in the core during reactor operation. 
\begin{figure}[ht!] % replace 't' with 'b' to force it to 
  \centering
  \includegraphics[width=\textwidth]{spectrum.png} 
  \caption{Neutron flux energy spectrum normalized by unit lethargy for initial and equilibrium fuel salt composition.}
  \label{fig:spectrum}
\end{figure}
Figure~\ref{fig:spectrum_zones} shows that zone I produced more thermal neutrons than zone II, corresponding to a majority of fissions occuring in the central part of the core. In the undermoderated zone II, the neutron energy spectrum is harder which leads to more capture of neutrons by $^{232}$Th and helps a achieve relatively high breeding ratio. Moreover, the (n,$\gamma$) resonance energy range in $^{232}$Th is from 10$^{-4}$ to 10$^{-2}$ MeV. Therefore, the moderator-to-fuel ratio for zone II was chosen to shift the neutron energy spectrum in this range. Furthermore, in the central core region (zone I), the neutron energy spectrum shifts to a harder spectrum over 20 years of reactor operation. In contrast, in the outer core region (zone II) a similar spectral shift takes place at a reduced scale. This resuls is in a good agreement with original ORNL report \cite{robertson_conceptual_1971} and most recent whole-core steady-state study \cite{park_whole_2015}.

@@

\subsection{Neutron flux}
Figure~\ref{fig:radial_flux} shows the radial distribution of fast and thermal neutron flux for both initial and equilibrium composition. The neutron flux has the same shape for both compositions but the equilibrium case has a harder spectrum. A significant spectral shift was observed for the central region of the core (zone I) when for the outer region (zone II) it is negligible for fast but notable for thermal neutrons. This neutron flux radial distribution is in a good agreement with original ORNL report \cite{robertson_conceptual_1971}. Overall, spectrum hardening during \gls{MSBR} operation should be carefully studied for designing the reactivity control system.
\begin{figure}[ht!] % replace 't' with 'b' to force it to 
  \includegraphics[width=\textwidth]{radial_flux.png} 
  \caption{Radial neutron flux distribution for initial and equilibrium fuel salt composition.}
  \label{fig:radial_flux}

@@

Table~\ref{tab:powgen_fraction} shows the power fraction in each zone for initial and equilibrium fuel composition. Figure~\ref{fig:pow_den} reflects the normalized power distribution of the \gls{MSBR} quarter core which is the same at both states. For both the initial and equilibrium compositions, fission primarily occurs in the center of the core, namely zone I. The spectral shift during reactor operation results in different power fractions at startup and equilibrium, but most of the power is still generated in zone I. 

@@

Table~\ref{tab:tcoef} summarizes temperature effects on reactivity calculated in this work for both initial and equilibrium fuel composition, and compared with original \gls{ORNL} report data \cite{robertson_conceptual_1971}. Uncertainty for each temperature coefficient also appears in Table~\ref{tab:tcoef}. The main physical principle underlying the reactor temperature feedback is an expansion of matterial that is heated. When the fuel salt temperature increases, the density of the salt decreases, but at the same time, the total volume of fuel salt in the core remains constant because it is bounded by the graphite. When the graphite temperature increases, the density of graphite decreases creating additional space for fuel salt. To determine temperature coefficients, the cross section temperatures for fuel and moderator were changed from 900K to 1000K. Three different cases were considered:

@@

In the first case, changes in the fuel temperature only impact fuel density. In this case, the geometry is unchanged because the fuel is a liquid. However, when the moderator heats up, both the density and the geometry change due to thermal expansion of the solid graphite blocks and reflector. Accordingly, the new graphite density was calculated using a linear temperature expansion coefficient of 1.3$\times10^{-6}$1/K \cite{robertson_conceptual_1971}. A new geometry input was created based on this information.

The fuel temperature coefficient (FTC) is negative for both initial and equilibrium fuel compositon due to thermal Doppler broadening of the resonance capture cross sections in the thorium and is in a good agreement with earlier research \cite{robertson_conceptual_1971,park_whole_2015}. The moderator temperature coefficient (MTC) is positive for startup composition and decreases during reactor operation because of spectrum hardening with fuel depletion. Finally, the total temperature coefficient of reactivity is negative for both cases, but decreases during reactor operation due to spectral shift. In summary, even after 20 years of operation the total temperature coefficient of reactivity is relatively large and negative during reactor operation, despite positive MTC, and affords excellent reactor stability and control.

\subsection{Reactivity control system rod worth}
Table~\ref{tab:rod_worth} summarizes the reactivity control system worth. During normal operation the control (graphite) rods are fully inserted, and the safety (B$_4$C) rods are fully withdrawn. To insert negative reactivity into the core, the graphite rods are gradually withdrawn from the core. In an accident, the safety rods would fall down into the core. The integral rod worths were calculated for various positions to separately estimate control (graphite) rod, safety (B$_4$C) rod, and the whole reactivity control system worth. Control rod integral worth is approximately 28 cents and stays almost constant during reactor operation. The safety rod integral worth decreases by  16.2\% during 20 years of operation because of neutron spectrum hardening and absorber accumulation in proximity to reactivity control system rods. This 16\% decline in control system worth should be taken into account in \gls{MSBR} accident analysis and safety justification.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{table}[ht!]
  \centering
  \caption{Control system rod worth for initial and equilibrium fuel composition.}
\begin{tabularx}{\textwidth}{ b | x | x } \hline
Reactivity parameter [cents]  &  Initial      &  Equilibrium      \\ \hline
Control (graphite) rod integral worth               & $\ 28.2\pm0.8$    & $\ 29.0\pm0.8$ \\ 
Safety (B$_4$C) rod integral worth                  & $251.8\pm0.8$    & $211.0\pm0.8$  \\
Total reactivity control system worth               & $505.8\pm0.7$    & $424.9\pm0.8$ \\ \hline
\end{tabularx}
  \label{tab:rod_worth}
\end{table}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

>>>>

\section{Results}
The SaltProc online reprocessing simulation package is demonstrated in four 
applications: (1) analyzing  \gls{MSBR} neutronics and fuel cycle to find the 
equilibrium core composition and core depletion, (2) studying operational and 
safety parameters evolution during \gls{MSBR} operation, (3) demonstrating that 
in a single-fluid two-region \gls{MSBR} conceptual design the undermoderated 
outer core zone II works as a virtual ``blanket", reduces neutron leakage and 
improves breeding ratio due to neutron energy spectral shift, and (4) 
determining the effect of fission product removal on the core neutronics.

The neutron population per cycle and the number of active/inactive cycles were 
chosen to obtain balance between reasonable uncertainty for a transport problem 
($\leq$ 15 pcm\footnote{ 1 pcm = 10$^{-5}\Delta k_{eff}/k_{eff}$} for effective 
multiplication factor) and computational time. The \gls{MSBR} depletion and 
safety parameter computations were performed on 64 Blue Waters XK7 nodes (two 
AMD 6276 Interlagos CPU per node, 16 floating-point Bulldozer core units per 
node or 32 ``integer" cores per node, nominal clock speed is 2.45 GHz). The 
total computational time for calculating the equilibrium composition was 
approximately 9,900 node-hours (18 core-years.)

\subsection{Effective multiplication factor}
Figure~\ref{fig:keff} shows the effective multiplication factors 
obtained using SaltProc and SERPENT 2. The effective multiplication factors are 
calculated after removing fission products listed in 
Table~\ref{tab:reprocessing_list} and adding the fertile material at the end of 
cycle time\footnote{ The \gls{MSBR} program defined a ``cycle time" as the 
        amount of time required to remove 100\% of a target nuclide from a fuel 
salt.} which was fixed at 3 days for this work. The effective multiplication 
factor fluctuates significantly as a result of the batch-wise nature of this 
online reprocessing strategy.

\begin{figure}[ht!] 
  \centering
  \includegraphics[width=\textwidth]{keff.png}
  \caption{Effective multiplication factor dynamics for full-core \gls{MSBR} 
  model for a 60-year reactor operation. The confidence interval 
  $\sigma=\pm15pcm$ is shaded.}
  \label{fig:keff}
\end{figure}

First, SERPENT calculates the effective multiplication factor for the beginning 
of cycle time (fresh fuel composition for the first step). Next, it computes 
the new fuel salt composition for the end of a 3-day depletion step. The 
corresponding effective multiplication factor is much smaller than the previous 
one. Finally, SERPENT calculates $k_{eff}$ for the depleted composition after 
applying feeds and removals. $K_{eff}$ this increases accordingly since major reactor 
poisons (e.g. Xe, Kr) are removed, while fresh fissile material ($^{233}$U) 
from the protactinium decay tank is added.

Additionally, the presence of rubidium, strontium, cesium, and barium in the 
core are disadvantageous to reactor physics. In fact, removal of these elements 
every 3435 days causes the multiplication factor to jump by approximately 450 
pcm, and limits using the batch approach for online reprocessing simulation. 
Overall, the effective multiplication factor gradually decreases from 1.075 to 
$k_{eff} \approx 1.02$ at equilibrium after approximately 6 years of 
irradiation.

\subsection{Fuel salt composition dynamics}

The analysis of the fuel salt composition evolution provides more comprehensive 
information about the equilibrium state. Figure~\ref{fig:adens_eq} shows number 
density of major nuclides which have a strong influence on the reactor core 
physics. Concentration of $^{233}$U, $^{232}$Th, $^{233}$Pa, and $^{232}$Pa in 
fuel salt change insignificantly after approximately 2500 days of operation. 
Particularly, $^{233}$U number density fluctuates less than 0.8\% in the time 
interval from 16 to 20 years of operation, hence,a quasi-equilibrium state was 
achieved after 16 years of reactor operation.
\begin{figure}[ht!] % replace 't' with 'b' to 
  \centering
  \includegraphics[width=\textwidth]{major_isotopes_adens.png}
  \caption{Number density of major nuclides during 60 years of reactor 
  operation.}
  \label{fig:adens_eq}
\end{figure}
In contrast, a wide variety of nuclides, including fissile isotopes (e.g. 
$^{235}$U) and non-fissile strong absorbers (e.g. $^{234}$U), keep accumulating 
in the core. Figure~\ref{fig:fissile_short} demonstrates production of fissile 
isotopes in the core. In the end of the considered operational time, the core 
contains significant $^{235}$U ($\approx10^{-5}$ atom/b-cm), $^{239}$Pu 
($\approx5\times10^{-7}$ atom/b-cm), and $^{241}$Pu ($\approx 5\times10^{-7}$ 
atom/b-cm). Meanwhile, the equilibrium number density of the target fissile 
isotope $^{233}$U was approximately 7.97$\times10^{-5}$ atom/b-cm. Thus, 
production of new fissile materials in the core as well as $^{233}$U breeding 
make it possible to compensate for negative effects of strong absorber 
accumulation and keep the reactor critical.
\begin{figure}[htp!] % replace 't' with 'b' to \centering
  \includegraphics[width=\textwidth]{fissile_short.png}
  \caption{Number density of fissile in epithermal spectrum nuclides 
  accumulation during the reactor operation.}
  \label{fig:fissile_short}
\end{figure}

\subsection{Neutron spectrum}
Figure~\ref{fig:spectrum} shows the normalized neutron flux spectrum for the 
full-core \gls{MSBR} model in the energy range from $10^{-8}$ to $10$ MeV. The 
neutron energy spectrum at equilibrium is harder than at startup due to 
$^{238}$Pu, $^{239}$Pu, $^{240}$Pu, $^{241}$Pu, and $^{242}$Pu accumulation in 
the core during reactor operation.  \begin{figure}[ht!] % replace 't' with 'b' 
        to force it to \centering
  \includegraphics[width=\textwidth]{spectrum.png} \caption{Neutron flux energy 
  spectrum normalized by unit lethargy for initial and equilibrium fuel salt 
  composition.}
  \label{fig:spectrum}
\end{figure}

Figure~\ref{fig:spectrum_zones} shows that zone I produced more thermal neutrons 
than zone II, corresponding to a majority of fissions occuring in the central part 
of the core. In the undermoderated zone II, the neutron energy spectrum is harder 
which leads to more capture of neutrons by $^{232}$Th and helps a achieve relatively 
high breeding ratio. Moreover, the (n,$\gamma$) resonance energy range in $^{232}$Th 
is from 10$^{-4}$ to 10$^{-2}$ MeV. Therefore, the moderator-to-fuel ratio for zone 
II was chosen to shift the neutron energy spectrum in this range. Furthermore, in the 
central core region (zone I), the neutron energy spectrum shifts to a harder spectrum 
over 20 years of reactor operation. In contrast, in the outer core region (zone II) a 
similar spectral shift takes place at a reduced scale. This resuls is in a good 
agreement with original ORNL report \cite{robertson_conceptual_1971} and most recent 
whole-core steady-state study \cite{park_whole_2015}.

@@

\subsection{Neutron flux}
Figure~\ref{fig:radial_flux} shows the radial distribution of fast and thermal 
neutron flux for both initial and equilibrium composition. The neutron flux has 
the same shape for both compositions but the equilibrium case has a harder 
spectrum. A significant spectral shift was observed for the central region of 
the core (zone I) when for the outer region (zone II) it is negligible for fast 
but notable for thermal neutrons. This neutron flux radial distribution is in a 
good agreement with original ORNL report \cite{robertson_conceptual_1971}. 
Overall, spectrum hardening during \gls{MSBR} operation should be carefully 
studied for designing the reactivity control system.
\begin{figure}[ht!] % replace 't' with 'b' to force it to \centering
  \includegraphics[width=\textwidth]{radial_flux.png} \caption{Radial neutron 
  flux distribution for initial and equilibrium fuel salt composition.}
  \label{fig:radial_flux}
\end{figure}

@@

Table~\ref{tab:powgen_fraction} shows the power fraction in each zone for 
initial and equilibrium fuel composition. Figure~\ref{fig:pow_den} reflects the 
normalized power distribution of the \gls{MSBR} quarter core which is the same 
at both the initial and equilibrium states. For both the initial and equilibrium compositions, fission 
primarily occurs in the center of the core, namely zone I. The spectral shift 
during reactor operation results in different power fractions at startup and 
equilibrium, but most of the power is still generated in zone I at equilibrium. 

@@

Table~\ref{tab:tcoef} summarizes temperature effects on reactivity calculated 
in this work for both initial and equilibrium fuel composition, and compared 
with original \gls{ORNL} report data \cite{robertson_conceptual_1971}. 
Uncertainty for each temperature coefficient also appears in 
Table~\ref{tab:tcoef}. The main physical principle underlying the reactor 
temperature feedback is an expansion of material that is heated. When the fuel 
salt temperature increases, the density of the salt decreases, but at the same 
time, the total volume of fuel salt in the core remains constant because it is 
bounded by the graphite. When the graphite temperature increases, the density 
of graphite decreases creating additional space for fuel salt. To determine 
temperature coefficients, the cross section temperatures for fuel and moderator 
were changed from 900K to 1000K. Three different cases were considered:

@@

In the first case, changes in the fuel temperature only impact fuel density. In 
this case, the geometry is unchanged because the fuel is a liquid. However, 
when the moderator heats up, both the density and the geometry change due to 
thermal expansion of the solid graphite blocks and reflector. Accordingly, the 
new graphite density was calculated using a linear temperature expansion 
coefficient of 1.3$\times10^{-6}$1/K \cite{robertson_conceptual_1971}. A new 
geometry input was created based on this information.

The fuel temperature coefficient (FTC) is negative for both initial and 
equilibrium fuel composition due to thermal Doppler broadening of the resonance 
capture cross sections in the thorium and is in a good agreement with earlier 
research \cite{robertson_conceptual_1971,park_whole_2015}. The moderator 
temperature coefficient (MTC) is positive for startup composition and decreases 
during reactor operation because of spectrum hardening with fuel depletion. 
Finally, the total temperature coefficient of reactivity is negative for both 
cases, but decreases during reactor operation due to spectral shift. In 
summary, even after 20 years of operation the total temperature coefficient of 
reactivity is relatively large and negative during reactor operation, despite 
positive MTC, and affords excellent reactor stability and control.

\subsection{Reactivity control system rod worth}
Table~\ref{tab:rod_worth} summarizes the reactivity control system worth. 
During normal operation the control (graphite) rods are fully inserted, and the 
safety (B$_4$C) rods are fully withdrawn. To insert negative reactivity into 
the core, the graphite rods are gradually withdrawn from the core. In an 
accident, the safety rods would fall down into the core. The integral rod 
worths were calculated for various positions to separately estimate control 
graphite moderator rods\footnote{In \cite{robertson_conceptual_1971}, the 
graphite rods are referred to as ``control'' rods.}, safety (B$_4$C) rod, and 
the whole reactivity control system worth. Control rod integral worth is 
approximately 28 cents and stays almost constant during reactor operation. The 
safety rod integral worth decreases by  16.2\% during 20 years of operation 
because of neutron spectrum hardening and absorber accumulation in proximity to 
reactivity control system rods. This 16\% decline in control system worth 
should be taken into account in \gls{MSBR} accident analysis and safety 
justification.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{table}[ht!]
  \centering
  \caption{Control system rod worth for initial and equilibrium fuel 
  composition.}
\begin{tabularx}{\textwidth}{ b | x | x } \hline
Reactivity parameter [cents]  &  Initial      &  Equilibrium      \\ \hline
Control (graphite) rod integral worth               & $\ 28.2\pm0.8$    & $\ 
        29.0\pm0.8$ \\ Safety (B$_4$C) rod integral worth                  & 
        $251.8\pm0.8$    & $211.0\pm0.8$  \\
Total reactivity control system worth               & $505.8\pm0.7$    & 
        $424.9\pm0.8$ \\ \hline
\end{tabularx}
  \label{tab:rod_worth}
\end{table}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
